### 2021-05-02 doc/VerticalArrangements.tex
finished drafts of simulation sections as well as conclusion

As for wholesaler surplus, which appears in the third panel, the effects seen there are largely the reflection of those for retailers: wholesaler surplus increases in about 77\% of all upstream simulated mergers, 23\% of simulated vertical mergers, 20\% of simulated downstream mergers, and abput 15\% of simulated integrated mergers. %\textbf{comparing these results to \citeN{ST2020}, the effects are less stark (i.e. less of a transfer.}
However, only wholesaler surplus from upstream mergers first-order stochastically dominates the wholesaler surplus of the other merger types.

@@

Starting with vertical mergers, absent incumbent integration (0 integrated firms), about 36\% of vertical mergers benefit consumers and about 23\% of vertical mergers are net beneficial (i.e. increase total surplus).  The presence of a single 3rd party integrated firm (1 integrated firm) decreases the percentage of mergers that benefit consumers to 25\% and the percentage of mergers that increase  total surplus to 16\%, but leads median harm largely unchanged. Interestingly, while adding a second integrated incumbent to the market also lowers the percentage of beneficial mergers, adding a third integrated firm increases the percentage of mergers that benefits consumers to about 27\% but does not increase the percentage of net beneficial mergers. Adding a fourth integrated incumbent increases the percentage of mergers that benefit consumers to 32\% but decreases the percentage of netbeneficial mergers to about 11\%. Taken together, Figure \ref{fig:CVvertincumbBW_updownBW} indicates that the presence of incumbent integrated firms is harmful, and that the harm does not meaningfully lessen as additional integrated incumbents are included.

@@

Turning to upstream mergers, when no incumbent firms are in the market (0 integrated firms), there is a negative relationship between bargaining power and both consumer and total harm.  Median consumer harm decreases from 2.7\% of pre-merger expenditures when retailers have relatively more bargaining power to about 31\% of pre-merger expenditures when wholesalers have relatively more bargaining power. Likewise, median total harm increases from 0.7\% of pre-merger expenditures when retailers have relatively more bargaining power to 14\% when wholesalers have relatively more bargaining power. By contrast, for mergers between an integrated and unintegrated upstream supplier (1 integrated firms), there is a positive relationship between bargaining power and both consumer and total surplus. Median consumer harm decreases from about 8.5\% of pre-merger expenditures when retailers have relatively more bargaining power to about -20\% of pre-merger expenditures when wholesalers have relatively more bargaining power. Likewise, median total harm increases from about 3.5\% of pre-merger expenditures when retailers have relatively more bargaining power to -4.4\% when wholesalers have relatively more bargaining power. Adding 2 additional incumbent integrated firms (4 integrated firms only strengthens the positive relationship between bargaining power and harm.

For downstream mergers, when no incumbent firms are in the market (0 integrated firms), there is a positive relationship between bargaining power and both consumer and total harm. Median consumer harm decreases from about 18\% of pre-merger expenditures when retailers have relatively more bargaining power to about 1\% of pre-merger expenditures when wholesalers have relatively more bargaining power. Likewise, median total harm decreases from 7.5\% of pre-merger expenditures when retailers have relatively more bargaining power to 2\% when wholesalers have relatively more bargaining power. By contrast, mergers between an integrated and unintegrated upstream supplier (1 integrated firms), causes the box and whisker plots to rotate clockwise around equal bargaining power, strengthening the positive relationship between bargaining power and both consumer and total surplus.  Increasing wholesaler bargaining power causes median consumer harm to fall from 25\% of pre-merger revenues to -11\% of pre-merger revenues, while median total harm falls from 12.5\% to 2\% of pre-merger revenues. Adding additional rival incumbents (4 integrated firms) has the reverse effect, rotating the box and whisker plots counter-clockwise and weakening the still positive relationship between bargaining power and surplus. Here, increasing wholesaler bargaining power causes median consumer harm to fall from 14\% of pre-merger revenues to -1.7\% of pre-merger revenues, while median total harm falls from 5.3\% to 1.9\% of pre-merger revenues.

@@

In terms of the relationship between bargaining power and harm, integrated mergers are perhaps most similar to vertical mergers. Like vertical mergers, there is a strong monotonic relationship between bargaining power and harm. Median consumer harm increases from about 26\% of pre-merger expenditures when retailers have relatively more bargaining power to about -1.4\% of pre-merger expenditures when wholesalers have relatively more bargaining power. Likewise, median total harm increases from 16\% of pre-merger expenditures when retailers have relatively more bargaining power to -13\% when wholesalers have relatively more bargaining power.

@@

Here, we explore the role that the constant marginal cost assumption plays in driving merger outcomes. In theory, constant marginal costs can increase the magnitude of EDM, as the wholesaler price reduction applies equally to all units produced by the retailer. By contrast, if either the integrated wholesaler or retailer face linear marginal costs, increased output from the EDM is limited by the increased marginal costs, which should limit the benefit to the parties. Likewise, in theory, RRC could be more profitable for the merging parties under linear than constant marginal costs, as linear marginal costs cause the integrated wholesaler's price to increase more rapidly.

Because EDM is plausibly greater under constant marginal costs while RRC is also plausibly less, then one might expect that on net mergers are more beneficial under constant than linear marginal costs. Figure \ref{fig:surplussumcost} displays box and whisker plots that depict the net effect of these different marginal cost specifications on consumer and total welfare under six different scenarios:
linear marginal costs for the non-merging parties products (``Party Constant", dark blue), constant marginal costs for all products (``Constant", orange), linear marginal costs for retailers (``Constant/Linear", light grey), linear marginal costs for wholesalers (``Linear/Constant", dark grey), linear marginal costs for the merging parties products (``Party Linear",light blue) and linear marginal costs for all products (``Linear'',red). These six scenarios are ordered from least to most consumer harm in vertical mergers.

Overall, consumer harm is less prevalent for mergers where firms have constant marginal costs then when firms have linear costs. Mergers with all constant-cost firms (orange) benefit consumers in about 15\% of all mergers, while mergers with all linear-cost firms benefit consumers in only 2.3\% of mergers. Indeed, consumer harm under the all linear marginal costs scenario first-order stochastically dominates consumer harm under the all constant marginal costs scenario. The median consumer harm from markets with linear marginal costs is about 9.3\% of pre-merger total expenditures, about 1.3 times the magnitude of markets with constant marginal costs.

The biggest differences in outcomes occurs for vertical mergers. when all firms have constant marginal costs benefit consumers in about 30\% of all simulations , but only in about 2.3\% of simulations when all firms have linear costs (light blue). Upstream mergers with all constant marginal cost firms benefit consumers in about 12\% of simulations, but only 0.2\% of simulations when firms have linear costs. Downstream mergers with constant marginal cost firms benefit consumers in about 11\% of simulations, but only about 6\% of simulations when firms have linear costs. Finally, integrated  mergers benefit consumers in about 10\% of all simulations when firms have constant costs, but less than 1\% of simulations when firms have linear costs.

@@

The U.S. waste and recycling industry generates approximately \$80 billion in annual revenues. In recent years the industry has experienced significant merger activity including several large acquisitions between vertically integrated, national competitors. However, solid waste companies tend not to be homogenous in their degree of vertical integration across geographies. As a result, each of these mergers exhibits a variety of vertical supply chain configurations both pre- and post-merger across their relevent local markets, making it an excellent application for further study of the welfare impacts of mergers that have both horizontal and vertical aspects.

@@

In 2020, Waste Management, the largest waste management company in the U.S., acquired Advanced Disposal Services (ADS), previously the fourth largest company, for \$4.6 billion. GFL Environmental also acquired WCA Waste Corporation for \$1.2 billion in 2020. Republic Services, the second largest waste management company in the U.S., obtained Department of Justice (DOJ) approval to acquire Santek Environmental in March 2021. These three merged companies along with Waste Connections, the third largest waste management company, are estimated to contol over 60\% of available landfill capacity nationally and also rank among the top haulers nationwide.\footnote{\tiny{Waste Business Journal, \url{https://www.wastedive.com/news/public-companies-increased-control-of-74b-us-waste-industry-in-2018/556079/ }}} Concentration in local markets varies substantially, however, in both the upstream and downstream markets.

Concerns about vertical competitive effects were raised in both the Republic-Santek and the Waste Management-ADS transactions. Both the Solid Waste Agency of Lake County, IL and the Solid Waste Agency of Northern Cook County submitted comments in opposition to the proposed asset divestitutre from Waste Management-ADS, stating that a vertically integrated competitor was needed to maintain competition in their local market post-merger.\footnote{\tiny{See  Comments by SWALCO and SWANCC : U.S. and Plaintiff States v. Waste Management, Inc., and Advanced Disposal Services, Inc., \url{https://www.justice.gov/atr/case-document/file/1377646/download}.}} In the DOJ complaint filed in the Republic-Santek case, vertical competitive effects were alleged to arise from the combination of their integrated assets in the Chattanooga area.\footnote{\tiny{See U.S. and State of Alabama v. Republic Services, Inc. and Santek Waste Services, LLC, \url{https://www.justice.gov/atr/case-document/file/1382031/download}}}

Our application in progress further identifies local markets in which these acquisitions result in: 1) only horizontal combinations of assets, 2) only vertical combinations of assets, and 3) combinations of vertically integrated assets - in the presence of differing numbers of existing integrated competitors to analyze the wefare effects from mergers with complex vertical arrangements.

>>>>

As for wholesaler surplus, which appears in the third panel, the effects seen there are largely the reflection of those for retailers: wholesaler surplus increases in about 77\% of all upstream simulated mergers, 23\% of simulated vertical mergers, 20\% of simulated downstream mergers, and about 15\% of simulated integrated mergers. %\textbf{comparing these results to \citeN{ST2020}, the effects are less stark (i.e. less of a transfer.}
However, only wholesaler surplus from upstream mergers first-order stochastically dominates the wholesaler surplus of the other merger types.

@@

Starting with vertical mergers, absent incumbent integration (0 integrated firms), about 36\% of vertical mergers benefit consumers and about 23\% of vertical mergers are net beneficial (i.e. increase total surplus).  The presence of a single 3rd party integrated firm (1 integrated firm) decreases the percentage of mergers that benefit consumers to 25\% and the percentage of mergers that increase  total surplus to 16\%, but leads median harm largely unchanged. Interestingly, while adding a second integrated incumbent to the market also lowers the percentage of beneficial mergers, adding a third integrated firm increases the percentage of mergers that benefits consumers to about 27\% but does not increase the percentage of net beneficial mergers. Adding a fourth integrated incumbent increases the percentage of mergers that benefit consumers to 32\% but decreases the percentage of net beneficial mergers to about 11\%. Taken together, Figure \ref{fig:CVvertincumbBW_updownBW} indicates that the presence of incumbent integrated firms is harmful, and that the harm does not meaningfully lessen as additional integrated incumbents are included.

@@

Turning to upstream mergers, when no incumbent firms are in the market (0 integrated firms), there is a positive relationship between bargaining power and both consumer and total harm.  Median consumer harm increases from 2.2\% of pre-merger expenditures when retailers have relatively more bargaining power to about 29\% of pre-merger expenditures when wholesalers have relatively more bargaining power. Likewise, median total harm increases from 0.7\% of pre-merger expenditures when retailers have relatively more bargaining power to about 12\% when wholesalers have relatively more bargaining power. By contrast, for mergers between an integrated and unintegrated upstream supplier (1 integrated firms), there is a positive relationship between bargaining power and both consumer and total surplus. Median consumer harm decreases from about 8.4\% of pre-merger expenditures when retailers have relatively more bargaining power to about -24\% of pre-merger expenditures when wholesalers have relatively more bargaining power. Likewise, median total harm increases from about 3.5\% of pre-merger expenditures when retailers have relatively more bargaining power to -8.4\% when wholesalers have relatively more bargaining power. Adding 2 additional incumbent integrated firms (4 integrated firms) mitigates some of the adverse effects of mergers while still preserving the positive relationship between bargaining power and harm.

For downstream mergers, when no incumbent firms are in the market (0 integrated firms), there is a positive relationship between bargaining power and both consumer and total harm. Median consumer harm decreases from about 18\% of pre-merger expenditures when retailers have relatively more bargaining power to about 1\% of pre-merger expenditures when wholesalers have relatively more bargaining power. Likewise, median total harm decreases from 7.6\% of pre-merger expenditures when retailers have relatively more bargaining power to about 2\% when wholesalers have relatively more bargaining power. By contrast, mergers between an integrated and unintegrated upstream supplier (1 integrated firms), causes the box and whisker plots to rotate clockwise around equal bargaining power, strengthening the positive relationship between bargaining power and both consumer and total surplus.  Increasing wholesaler bargaining power causes median consumer harm to fall from 25\% of pre-merger revenues to -33\% of pre-merger revenues, while median total harm falls from 12.2\% to about 1\% of pre-merger revenues. Adding additional rival incumbents (4 integrated firms) has the reverse effect, rotating the box and whisker plots counter-clockwise and weakening the still positive relationship between bargaining power and surplus. Here, increasing wholesaler bargaining power causes median consumer harm to fall from 14\% of pre-merger revenues to -33\% of pre-merger revenues, while median total harm falls from 5.5\% to 3.99\% of pre-merger revenues.

@@

In terms of the relationship between bargaining power and harm, integrated mergers are perhaps most similar to vertical mergers. Like vertical mergers, there is a strong monotonic relationship between bargaining power and harm. Median consumer harm increases from about 26\% of pre-merger expenditures when retailers have relatively more bargaining power to about 3.75\% of pre-merger expenditures when wholesalers have relatively more bargaining power. Likewise, median total harm increases from 16\% of pre-merger expenditures when retailers have relatively more bargaining power to -14.5\% when wholesalers have relatively more bargaining power.

@@

Here, we explore the role that the constant marginal cost assumption plays in driving merger outcomes. In theory, constant marginal costs can increase the magnitude of EDM, as the wholesaler price reduction applies equally to all units produced by the retailer. By contrast, if either the integrated wholesaler or retailer face linear marginal costs, increased output from the EDM is limited by the increased marginal costs, which in turn limits the benefit to the parties. Likewise, in theory, RRC could be more profitable for the merging parties under linear than constant marginal costs, as linear marginal costs cause the integrated wholesaler's prices to increase more rapidly.

Because EDM is plausibly greater under constant marginal costs while RRC is also plausibly less, then one might expect that on net mergers are more beneficial under constant than linear marginal costs. We explore this hypothesis in Figure \ref{fig:surplussumcost}, which displays box and whisker plots that depict the net effect of these different marginal cost specifications on consumer and total welfare under four different scenarios:
constant marginal costs for all firms (top row, blue), linear marginal costs for all firms (top row, orange), constant marginal costs for the merging parties' products but linear costs for the non-merging parties products (bottom row, blue), and linear marginal costs for the merging parties' products but constant costs for the non-merging parties products (bottom row, orange).

Overall, harm is less prevalent for mergers where firms have constant marginal costs than where firms have linear costs. Mergers with constant-cost firms increases consumer surplus  in about 15\% of all mergers and total surplus in 10.5\% of all mergers, while mergers with linear-cost firms increase consumer surplus in only 2.3\% of mergers and total surplus in 0.8\% of all mergers. However, while the consumer and total harm distributions for each merger type under the all-firm linear marginal costs scenario first-order stochastically dominates their counterparts under the all-firm constant marginal costs scenario, the consumer and total harm distributions under the party linear marginal costs scenario only first-order stochastically dominates the their counterparts under the party constant marginal costs scenario for vertical and integrated mergers. The median consumer harm from markets with linear marginal costs is about 9.3\% of pre-merger total expenditures, about 1.3 times the magnitude of markets with constant marginal costs.

Changing the cost structure consistently yields the greatest differences in outcomes for vertical mergers. When all firms have constant marginal costs, vertical mergers increase consumer surplus in about 30\% of all vertical merger simulations and total surplus in about 18\% of all vertical merger simulations, but only increase consumer surplus in about 1\% of vertical merger simulations and total surplus in about 0.9\% of vertical merger simulations when all firms have linear costs. Moreover, qualitatively similar results hold for vertical mergers when only the merging parties' products have constant marginal costs, compared to when only the merging parties' products have linear costs.

The differences between the constant and linear cost structures is more muted for both upstream and downstream mergers. Upstream mergers with all constant marginal cost firms benefit consumers in about 12\% of simulations, but only 0.2\% of simulations when firms have linear costs. Downstream mergers with all constant marginal cost firms benefit consumers in about 11\% of simulations, but only about 6\% of simulations when all firms have linear costs. Again, qualitatively similar results hold for vertical mergers when only the merging parties' products have constant marginal costs, compared to when only the merging parties' products have linear costs.

Finally, while there is a marked difference in outcomes between the constant and linear cost structures for integrated  mergers under the all-firms scenario, the differences in outcomes are more muted under the party scenario, especially for consumer surplus. In particular, median consumer harm when all firms face constant marginal costs is 13.6\%, compared to the median consumer harm of 17.5\% when all firms face linear marginal costs. By contrast, median consumer harm when only the merging parties' have constant costs is 15.1\%, compared to the median consumer harm of 16.5\% when the parties' have linear costs.  %Integrated mergers benefit consumers in about 10\% of simulations when all firms have constant costs, but less than 1\% of simulations when all firms have linear costs.

@@

The U.S. waste and recycling industry generates approximately \$80 billion in annual revenues. In recent years the industry has experienced significant merger activity including several large acquisitions between vertically integrated, national competitors. However, solid waste companies tend not to be homogeneous in their degree of vertical integration across geographies. As a result, each of these mergers exhibits a variety of vertical supply chain configurations both pre- and post-merger across their relevant local markets, making it an excellent application for further study of the welfare impacts of mergers that have both horizontal and vertical aspects.

@@

In 2020, Waste Management, the largest waste management company in the U.S., acquired Advanced Disposal Services (ADS), previously the fourth largest company, for \$4.6 billion. GFL Environmental also acquired WCA Waste Corporation for \$1.2 billion in 2020. Republic Services, the second largest waste management company in the U.S., obtained Department of Justice (DOJ) approval to acquire Santek Environmental in March 2021. These three merged companies along with Waste Connections, the third largest waste management company, are estimated to control over 60\% of available landfill capacity nationally and also rank among the top haulers nationwide.\footnote{\tiny{Waste Business Journal, \url{https://www.wastedive.com/news/public-companies-increased-control-of-74b-us-waste-industry-in-2018/556079/ }}} Concentration in local markets varies substantially, however, in both the upstream and downstream markets.

Concerns about vertical competitive effects were raised in both the Republic-Santek and the Waste Management-ADS transactions. Both the Solid Waste Agency of Lake County, IL and the Solid Waste Agency of Northern Cook County submitted comments in opposition to the proposed asset divestiture from Waste Management-ADS, stating that a vertically integrated competitor was needed to maintain competition in their local market post-merger.\footnote{\tiny{See  Comments by SWALCO and SWANCC : U.S. and Plaintiff States v. Waste Management, Inc., and Advanced Disposal Services, Inc., \url{https://www.justice.gov/atr/case-document/file/1377646/download}.}} In the DOJ complaint filed in the Republic-Santek case, vertical competitive effects were alleged to arise from the combination of their integrated assets in the Chattanooga area.\footnote{\tiny{See U.S. and State of Alabama v. Republic Services, Inc. and Santek Waste Services, LLC, \url{https://www.justice.gov/atr/case-document/file/1382031/download}}}

Our application in progress further identifies local markets in which these acquisitions result in: 1) only horizontal combinations of assets, 2) only vertical combinations of assets, and 3) combinations of vertically integrated assets - in the presence of differing numbers of existing integrated competitors to analyze the welfare effects from mergers with complex vertical arrangements.

\section{Conclusion}

This paper relaxes two strong assumptions made in \citeN{ST2020}: that mergers occur in vertical supply chains where in the pre-merger state, no firms are vertically integrated, and that all firms employ constant marginal cost technology. Relaxing these constraints is important as competition authorities are called upon to investigate mergers where either the merging parties or a third party are already vertically integrated, or where the presence of a scarce input could reduce the benefit from EDM or increase the harm from RRC. Here, we show using both numerical simulations that relaxing these assumptions can result in merger outcomes that are markedly different from those when these assumptions are maintained. We also show using simulation that relative bargaining power can serve as a useful indicia for predicting consumer and total harm from mergers where these two assumptions are relaxed. Finally, we intend to use two recent trash mergers to investigate the importance of these assumptions in predicting merger harm. Our hope is that future research will focus on some of the other strong assumptions made in \citeN{ST2020}, such as how the presence of either two-part tariffs or large fixed costs affect the Nash bargaining game and therefore merger outcomes.
